Formulating the posterior for \tardis

### 2015-08-20 posterior/model.tex
[posterior] Major overhaul of likelihood description
* discuss general and large-n separately
* add mathematica notebook to show what I tried, what worked and what
  didn't

@@

@@

The first term in \refeq{lum} gives the luminosity in bin $i$ given
$N_i$ packets, so $\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$
is the luminosity of packet $j$ in bin $i$. Obviously $P(\Lumi | 0,
\vecth) = \delta(\Lumi)$. The crucial term is
$P\given{L_i}{N_i=1,\vecth}$: it is the distribution of luminosity of
a single packet in bin $i$ and determines all other terms with
$N_i>1$.

We thought hard and found it a fair assumption that the distribution
of $l_{ij}$ is actually independent of the frequency thus independent
of $i$ and $j$ and we can drop the subscripts for simplicity, $l_{ij}
\to r$. The reason is that each packet initially gets the same
luminosity and so far it seems the amount of luminosity it gains or
loses is independent of the frequency. The question then is what is
the functional form $P(r | \vecth) \equiv P(\Lumi | N_i=1, \vecth)
$. The distributions $P(\Lumi | N_i, \vecth)$ with $N_i > 1$ can be
determined from the case of $N_i=1$ using convolutions for small $N_i$
and the central limit theorem for $N_i \ge 15$.

An approximation of $P(r | \vecth)$---a histogram of the luminosities
of all packets that leave the supernova---is shown in
\reffig{hist}.
\begin{figure*}[h]
  \begin{center}
    \includegraphics[width=\textwidth]{fit}
  \end{center}
  \caption{
  }
  \label{fig:hist}
\end{figure*}
The prominent features are a long tail at low luminosity, a sharp
peak, and a clear cutoff at high luminosity.  After researching quite
a bit, I found a flexible unimodal univariate distribution that can
accomodate this skewed distribution, it is the generalized \betap{}
distribution that depends on the hyperparameters $\vecnu \equiv (a, s,
\alpha, \gamma, \beta)$
\begin{equation}
  \label{eq:betaprime}
  P(r | \vecth) = \genbetapr(r | \vecnu(\vecth)) \equiv \frac{1}{B(\alpha, \gamma)} \left|\frac{\beta}{s} \right| \left( \frac{r - a}{s} \right)^{\alpha \beta - 1} \left(1+\left( \frac{r - a}{s} \right)^{\beta} \right)^{-\alpha - \gamma} \,.
\end{equation}

Since we assume it is the same in all bins we can then use the entire
set of samples that depend on $\vecth$ to infer $\vecnu | \vecth$. For
simplicity, we could use a maximum-likelihood fit and simply plug in
the result neglecting the uncertainty on $\vecnu$. This should be OK
as we usually have many packets; i.e., $N \gg 1$.

With $P(r|\vecth)$ fixed, we can now look at sums of luminosities. The
standard result is that if $r_1, r_2 \sim P(r|\vecth)$ then
\begin{equation}
  \label{eq:convolution}
  z_2 = r_1 + r_2 \sim \int \rmdx{r} P(r|\vecth) P(z_2 - r|\vecth) \,.
\end{equation}
Sums of $N_i$ terms are computed in identical fashion by convolution
of $P(r|\vecth)$ with the distribution of the sum with $N_i-1$
terms. For $N_i \gtrsim 15$, the central limit theorem applies and
\begin{equation}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \left(\Lumi | N_i * E_{P(r|\vecth)}[r], \sqrt{N_i * V_{P(r|\vecth)}[r]} \right)
\end{equation}
Note that for the convolutions, we would have to use the approximate
\genbetapr{} distribution but for the mean and std. deviation in
\refeq{clt}, we should use the sample mean and std. deviation to
bypass the extra uncertainty induced by assuming $P(r | \vecth) =
\genbetapr(r | \vecnu(\vecth))$ and fixed $\vecnu$.

@@

Let us now consider $P(N_i | \vecth)$, the second term in
\eqref{eq:lum}. As stated before, we assume the number of packets in
the given simulation follows a Poisson distribution whose rate
parameter $\lambda_i = \lambda_i(\vecth)$ has to be inferred from the
observed $n_i$. We can use the law of total probability
\begin{equation}
  P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
\end{equation}
\todo{postpone the integral over $\lambda$ then use result for
  compound Poisson distr. and finally integrate. Hopefully this can
  still all be done analytically} and Bayes' theorem
\begin{equation}
  P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
\end{equation}
If we choose Jeffrey's prior $P_0(\lambda_i) \propto
1/\sqrt{\lambda_i}$ then we get the analytical
result~\cite[Eq. 5]{Aggarwal:2011aa}
\begin{equation}
  P\given{N_i}{n_i} = \frac{n_i! [2 (N_i + n_i)]!}{4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}
\end{equation}
where $n_i = n_i(\vecth)$ is the number of packets in bin $i$ actually
observed in the simulation at hand.  In essence, $P(N_i | \vecth)$ is
the posterior predictive distribution to have $N_i$ packets given that
the simulation spit out $n_i(\vecth)$ packages. So $n_i$ is known,
while $N_i$ is a random variable and we take into account its
uncertainty by summing over $N_i$ in \refeq{lum}. This is relevant for bins where $n_i$ is small. If we got
2 packets, it is not so unlikely we could have gotten one package. But
of course with only half the packages, the expectation for the sum of
luminosities changes significantly.

@@

The polychord program only performs sampling and integration on the
unit hypercube with coordinates $\vecu$. Furthermore, it assumes that
the prior in these coordinates is uniform
\begin{equation}
  \label{eq:hypercube-prior}
  P(\vecu) = 1.
\end{equation}
In each iteration, we have to return the log likelihood for given
$\vecu$ passed in by polychord.  As above, suppose $\vecth$ represents
the parameters that are fed into the tardis simulation, then we need
to find the transform $\vecth(\vecu)$. The prior transforms as
\begin{equation}
  \label{eq:par-trafo}
  P(\vecth) \rmdx{\vecth} =  P(\vecu) \rmdx{\vecu},
\end{equation}
so with the flat prior
\begin{equation}
  \label{eq:par-trafo-likelihood}
  P(\vecth) = \left| \frac{\partial \theta_i}{\partial u_j} \right|^{-1} .
\end{equation}
There are two options:
\begin{enumerate}
\item \label{direct-transform} We can find $\vecth(\vecu)$ such that
  the Jacobian on the right in \refeq{par-trafo-likelihood} is the
  prior we want. The we just compute $\log L(\vecth(\vecu))$ for every
  sample $\vecu$ polychord provides.
\item If \ref{direct-transform} is not possible, we need to divide
  out the Jacobian and multiply by the actual prior we want, so for
  polychord we pretend we have a uniform prior and an \emph{effective} likelihood
  \begin{equation}
    \label{eq:eff-likelihood}
    L'(\vecu) = L(\vecth(\vecu)) P(\vecth(\vecu)) \left| \frac{\partial \theta_i}{\partial u_j} \right|
  \end{equation}
\end{enumerate}
Nested sampling is most efficient if the prior is close to the
posterior as measured by the relative entropy or Kullback-Leibler
divergence. \todo{cite Skilling} We therefore prefer
option~\ref{direct-transform}. While it restricts us to simple forms
of priors: uniform, Gaussian, or Dirichlet, this, fortunately, is
enough for our purposes. The explicit transformation is given below,
separately for each block of parameters.

@@

Suppose we have $K$ abundances subject to the constraints
\begin{align}
  \label{eq:unit-simplex}
  & 0 \leq \theta_k \leq 1, & \sum_{k}\theta_k = 1.
\end{align}
In other words, the abundances are defined on the unit simplex.  As
shown in \todo{cite ``Cruising the simplex''}, we can directly
transform $\vecu$ with $\dim \vecu = K$ to $\vecth$ as follows. First
create $K$ Gamma variates with the inverse-transform method, in python
code $\xi_k = \texttt{scipy.stats.gamma.ppf}(u_k, \alpha_k)$ with the
shape parameter $\alpha_k$ explained below. Then normalize the samples
as
\begin{equation}
  \label{eq:dirichlet-gamma}
  \theta_k = \frac{\xi_k}{\sum_{k'} \xi_{k'}}.
\end{equation}
such that
\begin{equation}
  \label{eq:dirichlet-prior}
  P(\vecth) = \Dirichlet(\vecth | \alpha)
\end{equation}

@@

The canonical choice of a prior for fractions such as relative
abundances is the Dirichlet
distribution~\cite[App.~B]{bishop_pattern_2006}
\begin{equation}
  \label{eq:dirichlet-def}
  \Dirichlet \given{\vecth}{\vecalpha} = \frac{\Gamma\left(\sum_k \alpha_k\right)}{\prod_k \Gamma(\alpha_k)} \prod_{k=1}^K \theta_k^{\alpha_k-1}
\end{equation}
and depends on $K$ parameters $\vecalpha$. How to choose $\alpha_k$?
The mode of $\theta_k$ is $\frac{\alpha_k - 1}{\sum_k (\alpha_k -
  1)}$, so we should choose $\alpha_k > 1$ if we want a well defined
mode. For prior ignorance, we want $\alpha_k$ identical. The canonical
choice of an uninformative distribution is a uniform distribution on
the unit simplex given by $\alpha_k=1 \, \forall \, k$~\todo{cite
  posterior predictive hints}.

>>>>

\DeclareMathOperator{\erf}{erf}
\newcommand{\mui}{\mu_i}
% \newcommand{\Ni}{N_i}
% \newcommand{\ni}{n_i}
\newcommand{\sigi}{\sigma_i^2}
\DeclareMathOperator{\sgn}{sgn}
\newcommand{\El}{E[l]}
\newcommand{\Vl}{V[l]}

@@

All involved integrals are (assumed) absolutely convergent so we can
exchange the order of summation and integration, so we can combine
\refeq{like-expanded} and \refeq{lum} and we have to compute for each bin
\begin{align}
  \label{eq:sum-integral}
  \sum_{N_i=0}^{\infty} \int \rmdx{\Lumi} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}
  P(\Lumi | N_i, \vecth) P(N_i | \vecth).
\end{align}

@@

The first term in \refeq{lum}, $P(\Lumi | N_i, \vecth)$, gives the
probability of luminosity $\Lumi$ in bin $i$ given $N_i$ packets, so
$\Lumi = \sum_{j=1}^{N_i} \Lumi^j$ where $\Lumi^j$ is the luminosity
of packet $j$ in bin $i$. Obviously $P(\Lumi | 0, \vecth) =
\delta(\Lumi)$. The crucial term is
\begin{align}
  \label{eq:single-lum}
  P\given{\Lumi}{1, \vecth} \equiv P\given{L_i}{N_i=1,\vecth};
\end{align}
it is the distribution of luminosity of a single packet in bin $i$ and
determines all other terms with $N_i>1$ because $\forall j, L_i^j$ has
the same distribution $P\given{L_i}{1,\vecth}$.

For the real packets, it seems that $P(\Lumi | 1, \vecth)$ is
independent of the frequency bin $i$. A histogram is shown in
\reffig{hist}.  The prominent features are a long tail at low
luminosity, a sharp peak, and a clear cutoff at high luminosity.

\begin{figure*}[h]
  \begin{center}
    \includegraphics[width=\textwidth]{fit}
  \end{center}
  \caption{Distribution of luminosities of real packets.}
  \label{fig:hist}
\end{figure*}

For virtual packets, the luminosity distribution seems to depend
mildly on the frequency.  We do not know $P(\Lumi | 1, \vecth)$ before
we do the simulation and we do not know its functional form
either. But we want to minimize the assumptions on the form. Hence
seek a nonparametric description of the samples. Obviously, if we
observe $n_i=0$ or $n_i=1$, we cannot infer the distribution $P(\Lumi
| 1, \vecth)$ from such little information. If we may assume that
$P(\Lumi | 1, \vecth)$ varies only little with $i$, we may pool
together luminosities $l_{ij}, j=1 \dots n_i$, of adjacent bins to
form a handful of larger bins, say $10-100$, in which we assume the
distribution is constant and we have enough packets, at least $\sum_i
n_i \gtrsim 500$. The I propose we use Gaussian kernel density
estimation (KDE) to obtain a smooth estimate of $P(\Lumi | 1, \vecth)$
that we can evaluate at arbitray $\Lumi$. In python, the easiest way
to do this is via \texttt{scipy.stats.gaussian\_kde}. The fastest to
my knowledge is \texttt{figtree} but the difference will be mostly
irrelevant unless there $\sum_i n_i = 10^5$ packets in a combination
of bins.

Given the KDE estimate, we could compute the distribution for the sum
of two variables as the convolution
\begin{align}
  \label{eq:sum-of-2}
  P(\Lumi | N_i=2, \vecth) = \int \rmdx{l} P(\Lumi -l | N_i=1, \vecth) P(l | N_i=1, \vecth).
\end{align}
For $N_i =k$, we would repeat the procedure convolving $P(\cdot |
N_i=k-1, \vecth)$ and $P(\Lumi | N_i=1, \vecth)$ and each time we
would need a new KDE estimate and we would have to repeat the integral
for different values of $\Lumi$. This quickly gets out of hand and I
fear the errors due to KDE assumptions add up. In my attempts with
\texttt{mathematica}, the probability for many values of $\Lumi$ was
negative and the whole procedure is slow. Let's forget about it.

The more model-independent and faster approach is to use the
luminosity samples $\{l_j: j=1 \dots \sum n_i\}$ for all bins $i$ that
are combined. We can create samples of the sum of $N_i=2$ luminosities
by adding $l_1+l_2, l_3 + l_4 \dots$ and similarly for larger
$N_i$.

\subsubsection*{Approximation for large $n_i$}

For $N \gtrsim 15$, we can just use the central limit theorem
such that
\begin{align}
  \label{eq:clt}
  P(\Lumi | N_i, \vecth) \approx \gaussian \given{\Lumi}{N_i \times E[l], N_i \times V[l]}
\end{align}
and the expecation value and variance have to be estimated by the mean
and variance of the samples $\{l_j\}$. In fact, this paper~\cite{bohm_statistics_2014}
shows that the \emph{scaled Poisson} is a more accurate approximation
given the first and second moment. But the resulting integral can't be
solved exactly, so for easier calculation we stick with the Gaussian.

@@

Let us now consider $P(N_i | \vecth)$, the second term in
\eqref{eq:lum}. As stated before, we assume the number of packets in
the given simulation follows a Poisson distribution whose rate
parameter $\lambda_i = \lambda_i(\vecth)$ has to be inferred from the
observed $n_i = n_i(\vecth)$. We can use the law of total probability
\begin{equation}
  \label{eq:poisson-post-pred}
  P(N_i | \vecth) = \int \rmdx{\lambda_i} P(\lambda_i | \vecth) \poisson(N_i | \lambda_i, \vecth)
\end{equation}
and Bayes' theorem
\begin{equation}
  \label{eq:poisson-post}
  P(\lambda_i | \vecth) \propto \poisson(n_i | \lambda_i) P_0(\lambda_i)
\end{equation}
If we choose the noninformative and improper Jeffreys prior
$P_0(\lambda_i) \propto 1/\sqrt{\lambda_i}$ then we get the analytical
result~\cite[Eq. 5]{Aggarwal:2011aa}
\begin{align}
  \label{eq:post-pred-analytical}
  P\given{N_i}{\vecth} &= \frac{n_i! [2 (N_i + n_i)]!}{2^{N_i+n_i+1/2}4^{N_i} N_i! (2 n_i)! (N_i + n_i)!}\\
  &= \frac{(N_i + n_i + 1)_{N_i + n_i}}{(n_i + 1)_{n_i} \sqrt{2}\, 2^{3N_i + n_i}N_i!}
\end{align}
where
\begin{align}
  \label{eq:pochhammer}
  (a)_k \equiv a (a + 1) \dots (a + k - 1)
\end{align}
is the Pochhammer symbol.

In essence, $P(N_i | \vecth)$ is
the posterior predictive distribution to have $N_i$ packets given that
the simulation spit out $n_i(\vecth)$ packages. So $n_i$ is known,
while $N_i$ is a random variable and we take into account its
uncertainty by summing over $N_i$ in \refeq{lum}. This is relevant for bins where $n_i$ is small. If we got
2 packets, it is not so unlikely we could have gotten one package. But
of course with only half the packages, the expectation for the sum of
luminosities changes significantly.

\subsubsection*{Approximation for large $n_i$}

The mode and variance of $P(N_i | \vecth)$ are
\begin{align}
  \label{eq:poisson-mode-variance}
  \arg \max P(N_i | \vecth) = n_i -1, && V[N_i] = 2 n_i + 1.
\end{align}
Why is the variance $\sim 2 n_i$ and not $n_i$ as one would expect for Poisson? Variances are additive and we get one factor of $n_i$ from the Poisson in \refeq{poisson-post-pred} and another one from \refeq{poisson-post}.

For $n_i=0$, the distribution diverges at zero, a consequence of the
Jeffreys prior.  For $n_i \gtrsim 30$, the distribution can be well
approximated by a Gaussian
\begin{align}
  \label{eq:poisson-gauss-approx}
  P(N_i | \vecth) \approx \gaussian \given{N_i}{n_i-1, 2 n_i + 1}
\end{align}

\subsection{Putting it all together}

Now it is time to combine all the equations above to see what actually
needs to be implemented. The likelihood is
\begin{align}
  \label{eq:all-likelihood}
  P\given{\data}{\vecth} &= \prod_{i=1}^{N_b} \int \rmdx{\Lumi} P\given{\Lumi}{\vecth}   P\given{\data}{\Lumi}\\
  &= \prod_{i=0}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi} P \given{\Lumi}{N_i}
  P\given{N_i}{\vecth} P\given{\Lumi}{\mui, \sigi}
\end{align}

\subsubsection*{Large $n_i$}

If many packets are observed in bin $i$, then with good approximation
everything is Gaussian and we can evaluate the integral over $\Lumi$
analytically: the convolution of two Gaussians is another Gaussian
with variances added.

\begin{align}
  \label{eq:all-large-n}
P\given{\data}{\vecth}  &=  \prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi} \gaussian \given{N_i}{n_i-1, 2 n_i + 1} \gaussian \given{\Lumi}{N_i  E[l], N_i V[l]}\\
  & \phantom{\prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \int \rmdx{\Lumi}} \times \gaussian \given{\Lumi}{\mu_i, \sigma_i} \nonumber\\
  & = \prod_{i=1}^{N_b} \sum_{N_i=0}^\infty \gaussian \given{N_i}{n_i-1, 2 n_i + 1} \gaussian \given{N_i \El}{\mu_i, \sigi + N_i \Vl}
  % &= \frac{\left| \Vl \mui + \El \sigi \right|+ (\Vl \mui + \El \sigi) \erf\left(\frac{\sqrt{N_i} \left| \Vl \mui + \El \sigi \right|}{\sqrt{2 \sigi \Vl (\Vl N_i + \sigi)}}\right)}{}\\
%   &= \frac{1 + \sgn(\rho) \erf\left(\frac{\sqrt{N_i} \left| \rho \right|}{\sqrt{2 \sigi \Vl \xi}}\right)}{4 \pi \sqrt{(2n_i + 1) \xi}} \exp \left(-\frac{ (N_i \El -\mui)^2}{2  \xi}\right)
% \\
%   &\phantom{=} \times \exp \left(\frac{(2n_i+1) (N_i \El)^2 + N_i^3 \Vl - 2(2n_i+1)N_i \El \mui + (2n_i+1) \mu^2}{2 (2n_i + 1) \xi}\right) \nonumber\\
%   &\phantom{=} \times \exp \left(\frac{(n_i-1)^2 \sigi + (n_i - 1) N_i ((n_i-1)\Vl - 2\sigi) + N_i^2 (-2(n-1)\Vl + \sigi)}{2 (2n_i + 1) \xi}\right) \nonumber
\end{align}
% where
% \begin{align}
%   \label{eq:rho}
%   \rho &\equiv  \Vl \mui + \El \sigi\\
%   \xi &\equiv N_i \Vl  + \sigi
% \end{align}

For each bin, we have to compute a series. We have three options:
\begin{enumerate}
  \item The quick and dirty approach is to approximate $\sum_{N_i}$ by
  $\int \rmdx{N_i}$ and use \texttt{scipy.integrate.quad} to
  numerically integrate.
  \item If the first approach turns out to be too slow, we can just
  sum up a finite number of terms as the summand quickly falls off for
  large $N_i$; cf. \reffig{series}. The summand/integrand is unimodal
  and well behaved. For the numbers I tried, the mode was near $n_i$
  as implied by the first Gaussian. The mode can be computed
  analytically but the expression is very long, so I don't show it
  here. We could start summing from there and grow to the left and
  right until terms become negligibly small. In my tests in
  mathematica, I chose the number of terms by eye.
  \item The third and fastest way would be to compute the mode, then
  use the Laplace approximation to estimate the integral. This
  requires the 2nd derivative around the mode, which is available in
  closed form but too long to report here. Check the mathematica
  notebook for the explicit results.
\end{enumerate}
For tests I did, the first two agreed within 5 or 6 digits, and the
Laplace approximation good to the 3rd digit, so we might want to go
with that, as it should be the fastest.

For $N_i \to 0$, the summand is independent of $\El$ and $\Vl$ and
very small if $n_i$ is large, so there is fortunately no special case
to take care of.

\begin{figure}[h]
  \centering
  \includegraphics{series}
  \caption{Series summand $\gaussian \given{N_i}{n_i-1, 2 n_i + 1}
    \gaussian \given{N_i \El}{\mu_i, \sigi + N_i \Vl}$ of
    \refeq{all-large-n} for various values of $N_i$ and $\mui=5,
    \sigi=5^2,n_i=15, \El= 0.9 * \mui / n_i, \Vl=0.1$.}
  \label{fig:series}
\end{figure}

\subsubsection*{General case}

Plugging in the KDE estimates of $P\given{\Lumi}{N_i}$, we cannot use
the Laplace approximation because the mode and the derivative there
are not known. For small $n_i$, typically few terms in the series
contribute, so the number of numerical integrals can be kept
small. However, this depends on the experimental uncertainty. If it is
large, then we have to take into account more terms.

It is worth noting that for $N_i=0$, $\Lumi$ would be the sum of zero
terms and that is ill defined. So we have to start summing at
$N_i=1$. But that does not mean that $n_i=0$ is invalid, too. In fact,
$P(N_i=1 | n_i=0) = 17.7\%$ in our model with the Jeffreys prior. So
no packets seen means you could have seen a packet with some sizable
probability. Of course this is a simplification. Looking at a spectrum
in which the last 1000 bins didn't get a packet and the luminosity
drops before that, I associate a much smaller probability for $N_i=1$
in the very last bin than in bin $N_b-1000$. The difference is because
I take into account extra information: smoothness of the spectrum,
monotonicity \dots. At present, I don't know how to encode this in the
model \emph{and} keep it computationally tractable. It is left as an
exercise for the reader.

\subsubsection*{Recommendation}

The strategy would be to look at each bin separately. If $n_i \gtrsim
5$, use the large-$n_i$ approach from above. Else start at $N_i=1$ and
sum the numerical integrals until it is $10^{-4}$ of the maximum
term. For 10000 bins, this works can be parallelized easily.

For the real packets, I compared the general approach to the Gaussian
large-$n_i$ approximation, using numerical integration in both cases
in \texttt{mathematica}. For $n_i=1\dots5$, the difference is slighly
less than 10\%. From $n_i=10$ to 15, it drops from 5\% to 3\% level,
and for $n_i=100$, both approaches agree to within $10^{-4}$. Laplace
can be used from $n_i=5$ on. In all cases except $n_i=100$, the
Gaussian approximation is systematically \emph{below} the general
result.

If we are ok with less than 10\% error, we can employ the Gaussian
approximation from $n_i\ge5$, which means we neglect the 3rd and
higher moments of the luminosity distribution. If the virtual packets
are even more skewed than the real packets, we might need to start at
$n_i \sim 10$.

@@

The polychord program only performs sampling and integration on the
unit hypercube with coordinates $\vecu$. Furthermore, it assumes that
the prior in these coordinates is uniform
\begin{equation}
  \label{eq:hypercube-prior}
  P(\vecu) = 1.
\end{equation}
In each iteration, we have to return the log likelihood for given
$\vecu$ passed in by polychord.  As above, suppose $\vecth$ represents
the parameters that are fed into the tardis simulation, then we need
to find the transform $\vecth(\vecu)$. The prior transforms as
\begin{equation}
  \label{eq:par-trafo}
  P(\vecth) \rmdx{\vecth} =  P(\vecu) \rmdx{\vecu},
\end{equation}
so with the flat prior
\begin{equation}
  \label{eq:par-trafo-likelihood}
  P(\vecth) = \left| \frac{\partial \theta_i}{\partial u_j} \right|^{-1} .
\end{equation}
There are two options:
\begin{enumerate}
\item \label{direct-transform} We can find $\vecth(\vecu)$ such that
  the Jacobian on the right in \refeq{par-trafo-likelihood} is the
  prior we want. The we just compute $\log L(\vecth(\vecu))$ for every
  sample $\vecu$ polychord provides.
\item If \ref{direct-transform} is not possible, we need to divide
  out the Jacobian and multiply by the actual prior we want, so for
  polychord we pretend we have a uniform prior and an \emph{effective} likelihood
  \begin{equation}
    \label{eq:eff-likelihood}
    L'(\vecu) = L(\vecth(\vecu)) P(\vecth(\vecu)) \left| \frac{\partial \theta_i}{\partial u_j} \right|
  \end{equation}
\end{enumerate}
Nested sampling is most efficient if the prior is close to the
posterior as measured by the relative entropy or Kullback-Leibler
divergence~\cite{skilling_nested_2006}. We therefore prefer
option~\ref{direct-transform}. While it restricts us to simple forms
of priors: uniform, Gaussian, or Dirichlet, this, fortunately, is
enough for our purposes. The explicit transformation is given below,
separately for each block of parameters.

@@

Suppose we have $K$ abundances subject to the constraints
\begin{align}
  \label{eq:unit-simplex}
  & 0 \leq \theta_k \leq 1, & \sum_{k}\theta_k = 1.
\end{align}
In other words, the abundances are defined on the unit simplex.  As
shown in~\cite{betancourt_cruising_2012}, we can directly transform
$\vecu$ with $\dim \vecu = K$ to $\vecth$ as follows. First create $K$
Gamma variates with the inverse-transform method, in python code
$\xi_k = \texttt{scipy.stats.gamma.ppf}(u_k, \alpha_k)$ with the shape
parameter $\alpha_k$ explained below. Then normalize the samples as
\begin{equation}
  \label{eq:dirichlet-gamma}
  \theta_k = \frac{\xi_k}{\sum_{k'} \xi_{k'}}.
\end{equation}
such that
\begin{equation}
  \label{eq:dirichlet-prior}
  P(\vecth) = \Dirichlet(\vecth | \alpha)
\end{equation}

@@

The canonical choice of a prior for fractions such as relative
abundances is the Dirichlet
distribution~\cite[App.~B]{bishop_pattern_2006}
\begin{equation}
  \label{eq:dirichlet-def}
  \Dirichlet \given{\vecth}{\vecalpha} = \frac{\Gamma\left(\sum_k \alpha_k\right)}{\prod_k \Gamma(\alpha_k)} \prod_{k=1}^K \theta_k^{\alpha_k-1}
\end{equation}
and depends on $K$ parameters $\vecalpha$. How to choose $\alpha_k$?
The mode of $\theta_k$ is $\frac{\alpha_k - 1}{\sum_k (\alpha_k -
  1)}$, so we should choose $\alpha_k > 1$ if we want a well defined
mode. For prior ignorance, we want $\alpha_k$ identical. The canonical
choice of an uninformative distribution is a uniform distribution on
the unit simplex given by $\alpha_k=1 \, \forall \,
k$~\cite{tuyl_posterior_2009}.
